---
title: "Computational Statistics Computer Lab 2 (Group 7)"
author: 
  - Qinyuan Qi(qinqi464)
  - Satya Sai Naga Jaya Koushik	Pilla (satpi345)
date: "`r Sys.Date()`"
output: pdf_document
---

## Question 1:  Sampling algorithms for a triangle distribution (Solved by Qinqi Qi)

### Answer:

```{r setup1, include=FALSE}
########################## INIT CODE ###########################################
rm(list = ls())
```

### (A)
Since we need to find a envelope function for the density function, so we need
to plot the density function first. let $ x \in [-3, 3] $, and we know it's  
triangle sharped function. We also can plot it using the code(Check appendix).
the target function $g(x) = 1$ in $[-1,1]$, otherwise $g(x) = 0$

```{r 1a1 , echo=FALSE}
########################## [ 1 a ] #############################################

x <- seq(-2, 2, by = 0.001)
y <- rep(0, length(x))
z <- rep(0, length(x))

x1_index <- which(x > 1 | x < -1)
x2_index <- which(x >= -1 & x <= 0)
x3_index <- which(x > 0 & x <= 1)

y[x1_index] <- 0
y[x2_index] <- x[x2_index] + 1
y[x3_index] <- 1 - x[x3_index]

z0_index <- which(x > 1 | x < -1)
z1_index <- which(x >= -1 & x <= 1)

z[z1_index] <- 1
z[-z1_index] <- 0

data <- data.frame(x, y, z)

graph <- ggplot2::ggplot(data) +
  ggplot2::geom_line(mapping=ggplot2::aes(x=x,y=y)) + 
  ggplot2::geom_line(mapping=ggplot2::aes(x=x,y=z),color="blue") + 
  ggplot2::ggtitle("Density function of X and envelope function") +
  ggplot2::xlab("x") +
  ggplot2::ylab("y,z")

graph
```
After we get the plot, we can see that the density function is a triangle we 
will use an envelope function to cover the density function, and we can
use function $ g(x) = 1 $ to cover the density function in the [-1,1] range. 

Now we will write a code for the rejection sampling algorithm
our g function is $g(x) = 1$ in the range $[-1,1]$, otherwise $g(x) = 0$
our e function is $e(x) = 1 / a$ in the range $[-1,1]$, otherwise $e(x) = 0$
and we will use $a = 0.5$ here.

```{r 1a2}
generate_random_var_1a <- function(n, a) {
  samples <- numeric(n)
  # Generate a sample from the proposal distribution in [-1,1]
  count <- 1
  e <- 1 / a

  while (count <= n) {

    x <- runif(1, -1, 1)
    # Calculate the acceptance probability
    u <- runif(1)

    # since g will always be 1 in the range [-1,1], so y = 1
    # f1(Y) = 1 - Y and e(Y) = 1 / a = 2
    # f2(Y) = Y + 1 and e(Y) = 1 / a = 2
    # so we have f(Y) / e(Y) = (1 - Y) / 2 or Y+1 / e

    # Calculate the acceptance probability

    if (x > 0 && u <= (1 - x) / e) {
      samples[count] <- x
      count <- count + 1
    }

    if (x <= 0 && u <= (x + 1) / e) {
      samples[count] <- x
      count <- count + 1
    }
  }

  return(samples)
}

a <- 0.5
random_var_1a <- generate_random_var_1a(10000, a)
hist(random_var_1a)
```

### (B)

To generate a random variable from the density function using Inverse CDF,
we will need to calculate the CDF function first.

CDF function as follows:
$$

CDF = \left\{ \begin{array}
0 & \mbox{for} & x \in (-\inf,-1) \\
\frac{x^2}{2} + x & \mbox{for} & x \in (-1,0) \\
x - \frac{x^2}{2} & \mbox{for} & x \in (0,1) \\
1 & \mbox{for} & x \in (1,\inf) 
 \end{array}\right
$$

Since -Y has a triangle  distribution on $[-1,0]$, so we have 

$F^{-1}$ in range $[-1,0)$ is $x = -1 + \sqrt{1-2y}$,

Also we have $F^{-1}$ in $[0,1]$ range is $x =  1 - \sqrt{1-2y}$

The following is the function to generate a random variable from it

```{r 1b}
########################## [ 1 b ] #############################################
generate_random_var_1b <- function(n){
  u <- runif(n)
  x1 <-  - 1 + sqrt(1 - 2 * u)
  x2 <-  1 - sqrt(1 - 2 * u)
  return(c(x1, x2))
}

random_var_1b <- generate_random_var_1b(10000)
hist(random_var_1b)
```

### (C)
The code to generate a random variable following a triangle distribution as follows.

```{r 1c}
########################## [ 1 c ] #############################################
generate_random_var_1c <- function(n){
  u1_1c <- runif(n)
  u2_1c <- runif(n)
  return(u1_1c - u2_1c)
}

random_var_1c <- generate_random_var_1c(10000)
hist(random_var_1c)
```

### (D)

Since we already plot the data, we will not plot them here.

since the 1c is the simplest way to generate the random variable, but since it 
can not adapt to some specific distribution, for 1b , since we need to calculate 
CDF and inverse function, in some cases, it's hard to do that, so we will use 
1a to generate the random variable.


## Question 2:  Laplace distribution (Solved by Satya Sai Naga Jaya Koushik	Pilla)

### Answer:

```{r setup2, include=FALSE}
################### Init code for Question 1 ###################################
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
```
#### (a) 
```{r 2.a}
########################## [ 2 a ] ######################################
```

#### (b) 
```{r 2.b}
########################## [ 2 b ] #############################################
```

\newpage
# Appendix: Code for this report

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```

