---
title: "Computational Statistics Computer Lab 6 (Group 7)"
author: 
  - Qinyuan Qi(qinqi464)
  - Satya Sai Naga Jaya Koushik	Pilla (satpi345)
date: "`r Sys.Date()`"
output: pdf_document
---

## Question 1: Genetic algorithm (Solved by Qinyuan Qi)

### Answer:

### (1):

### (2):

### (3):

### (4):

### (5):

### (6):

### (7):

### (8):

### (9):

## Question 2: EM algorithm (Solved by Satya Sai Naga Jaya Koushik Pilla)

### Answer:

```{r setup2, include=FALSE}
########################## Init code for question 1 ############################
rm(list = ls())
library(ggplot2)
set.seed(12345)
```
### (1) Plot 2 histograms:

According to the plots generated, we found that the plot seems follow exponential distribution.

```{r 2.1}
################################## (2.1) #######################################
# Load data
data <- read.csv("censoredproc.csv", 
                sep = ";", header = TRUE)
# We will filter out the left-censored data which cens=2
filtered_data <- data[data$cens ==1,]
layout(matrix(c(1:2), 1, 2))
# plot the data
hist(data$time, breaks = 100, main="Hist of filtered data")
# plot the filtered data
hist(filtered_data$time, breaks = 100, main="Hist of filtered data")
```

### (2):

The general CDF form of an exponential distribution is:

$$
F(x,\lambda) = \begin{cases} 1 - e^{- \lambda x} & x \geq 0 \\ 0 & x < 0 \end{cases}
$$

So PDF of an exponential distribution is derivative of F on x:

$$
f(x,\lambda) = \begin{cases} \lambda e^{-\lambda x} & x \geq 0 \\ 0 & x < 0 \end{cases}
$$


Likelihood function for the exponential distribution is as follows.

$$
L(\lambda; x_{1},x_{2}...x_{n}) = \prod{f(x,\lambda)} = \lambda^{n} exp(-\lambda \sum_{j=1}^{n}x_{j})
$$

PDF for the truncated exponential distribution is derived as follows.

$$
P(X \leq x | X \leq c) = \frac{P(X \leq x, X \leq c)}{P(X \leq c)} = \frac{P(X \leq \lambda)}{P(X \leq c)} = \frac{\lambda e^{-\lambda x}}{ce^{-cx}} = \frac{\lambda}{c}e^{-\lambda x}
$$

So likelihood function for the truncated exponential distribution is as follows.

$$
L(\lambda|X \leq c; x_{1},x_{2}...x_{n}) = \prod{P(X \leq x | X \leq c)} = (\frac{\lambda^{n}}{c^{n}}) exp(-\lambda \sum_{j=1}^{n}x_{j})
$$

### (3):

Since it's relative straight forward , we will use the likelihood function directly.

So we will derive the EM function using the likelihood function in 2.2.

E-Step:

Let's compute the expectation of likelihood as follows.

$$
Q(\lambda,{\lambda}^{t}) = E(L(\lambda|X \leq c; x_{1},x_{2}...x_{n}),{\lambda}^{t})
$$

M-Step:

In M Step , we need to maximum Q with respect to  $\lambda$.

$$
\lambda^{t+1} = argmax_{\lambda} Q(\lambda,{\lambda}^{t})
$$

$$
Q(\lambda,{\lambda}^{t}) = E(L(\lambda|X \leq c; x_{1},x_{2}...x_{n}),{\lambda}^{t})
$$

### (4):

# Function to compute the E-step
estep <- function(lambda, x, c) {
  return(lambda / c * exp(-lambda * x))
}

# Function to compute the M-step
mstep <- function(lambda, x, c) {
  return(sum(x) / sum(c * exp(-lambda * x)))
}

# EM algorithm
em_algorithm <- function(initial_lambda, observed_data, truncation_point, max_iter = 100, tol = 1e-6) {
  lambda_current <- initial_lambda
  
  for (iter in 1:max_iter) {
    # E-step
    expected_values <- estep(lambda_current, observed_data, truncation_point)
    
    # M-step
    lambda_next <- mstep(lambda_current, observed_data, expected_values)
    
    # Check for convergence
    if (abs(lambda_next - lambda_current) < tol) {
      break
    }
    
    # Update lambda for the next iteration
    lambda_current <- lambda_next
  }
  
  return(list(lambda = lambda_current, iterations = iter))
}

# Example usage
set.seed(123)
observed_data <- rexp(100, rate = 0.5)  # Simulated truncated exponential data
truncation_point <- 2

# Initial guess for lambda
initial_lambda <- 0.5

# Run EM algorithm
result <- em_algorithm(initial_lambda, observed_data, truncation_point)

# Print the result
cat("Estimated lambda:", result$lambda, "\n")
cat("Number of iterations:", result$iterations, "\n")

### (5):

### (6):
